---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# burnr

<!-- badges: start -->
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/burnr)](https://cran.r-project.org/package=burnr)
[![Build Status](https://travis-ci.org/ltrr-arizona-edu/burnr.svg?branch=master)](https://travis-ci.org/ltrr-arizona-edu/burnr)
[![Coverage Status](https://coveralls.io/repos/github/ltrr-arizona-edu/burnr/badge.svg?branch=master)](https://coveralls.io/github/ltrr-arizona-edu/burnr?branch=master)
[![downloads](https://cranlogs.r-pkg.org/badges/burnr)](https://cran.r-project.org/package=burnr)
<!-- badges: end -->

Basic tools to analyze forest fire history data (e.g. FHX) in R. This is designed for power users and projects with special needs.

## Installation

You can install the released version of burnr from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("burnr")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ltrr-arizona-edu/burnr")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(burnr)

# This gives you a basic plot. There are more advanced options. For example, we can color our plot by sample species.

data(lgr2_meta)

plot(lgr2,
  color_group = lgr2_meta$SpeciesID,
  color_id = lgr2_meta$TreeID,
  plot_legend = TRUE
)
```

## Support

Documentation is included in the code. If you're new to `burnr`, our [2018 paper in Dendrochronologia](https://doi.org/10.1016/j.dendro.2018.02.005) is a nice survey of the package with many examples. We have also an [introduction](https://cran.r-project.org/package=burnr/vignettes/introduction.html). More information can be found on the [project's wiki](https://github.com/ltrr-arizona-edu/burnr/wiki). Note, this is still under construction.


## Citation

Please cite the original `burnr` paper if you use it in your research: 

    Malevich, Steven B., Christopher H. Guiterman, and Ellis Q. Margolis. 2018. 
    “Burnr: Fire History Analysis and Graphics in R.” Dendrochronologia 49 
    (June): 9–15. https://doi.org/10.1016/j.dendro.2018.02.005.


Citations help us to identify user needs and justify additional time developing and maintaining `burnr`.


## Development

Please file bugs in the [bug tracker](https://github.com/ltrr-arizona-edu/burnr/issues).

Want to contribute? Great! We're following [Hadley's packaging workflow](http://r-pkgs.had.co.nz/) and [style guide](http://style.tidyverse.org). Fork away.

If you're not a developer, don't worry! We also welcome help with documentation and tutorials.
